---
title: "World Health Organisation: Global Suicide Trends & Analysis"
author: "Chau dep trai"
output: 
  pdf_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1.1 Import & Tidy
7 countries removed (<= 3 years of data total)
2016 data was removed (few countries had any, those that did often had data missing)
HDI was removed due to 2/3 missing data
Generation variable has problems, detailed in 2.11
Continent was added to the dataset using the countrycode package
Africa has very few countries providing suicide data
```{r}
library(tidyverse) # general
library(ggalt) # dumbbell plots
library(countrycode) # continent
library(rworldmap) # quick country-level heat maps
library(gridExtra) # plots
library(broom) # significant trends within countries

theme_set(theme_light())
```

```{r}
# 1) Import & data cleaning

data_test <- read_csv("D:\\R\\RBAC\\RBAC\\master.csv")

glimpse(data)

# 2/3 missing value -> remove, not useable
sum(is.na(data$`HDI for year`)) / length(data$`HDI for year`)

# this variable is vague
table(data$age, data$generation)

data <- data %>%
  select(-c(`HDI for year`, `suicides/100k pop`)) %>%
  rename(gdp_for_year = `gdp_for_year ($)`,
         gdp_for_capita = `gdp_per_capita ($)`,
         country_year = `country-year`) %>%
  as.data.frame()

glimpse(data)
```
```{r}
# 2) OTHER ISSUES

# country_year has 12 rows for each country and year combination
data %>%
  group_by(country_year) %>%
  count() %>%
  filter(n != 12)

# missing of 2016 data
data <- data %>%
  filter(year != 2016) %>%
  select(-country_year)

# b) excluding countries with <= 3 years of data:
minimum_years <- data %>%
  group_by(country) %>%
  summarize(rows = n(), 
            years = rows / 12) %>%
  arrange(years)

data <- data %>%
  filter(!(country %in% head(minimum_years$country, 7)))

# no other major data issues found yet
```
```{r}
# 3) TIDYING DATAFRAME
glimpse(data)

data$age <- gsub(" years", "", data$age)
data$sex <- ifelse(data$sex == "male", "Male", "Female")

# getting continent data:
data$continent <- countrycode(sourcevar = data[, "country"],
                              origin = "country.name",
                              destination = "continent")

# Nominal factors
data_nominal <- c('country', 'sex', 'continent')
data[data_nominal] <- lapply(data[data_nominal], function(x){factor(x)})

# Making age ordinal
data$age <- factor(data$age, 
                   ordered = T, 
                   levels = c("5-14",
                              "15-24", 
                              "25-34", 
                              "35-54", 
                              "55-74", 
                              "75+"))

# Making generation ordinal
data$generation <- factor(data$generation, 
                   ordered = T, 
                   levels = c("G.I. Generation", 
                              "Silent",
                              "Boomers", 
                              "Generation X", 
                              "Millenials", 
                              "Generation Z"))
data <- as_tibble(data)
```
```{r}
# the global rate over the time period will be useful:

global_average <- (sum(as.numeric(data$suicides_no)) / sum(as.numeric(data$population))) * 100000
```
#1.2 Key Insights
Suicide rates are decreasing globally. (2.1)
Of those countries that show clear linear trends over time, 2/3 are decreasing. (2.5.2)
On average, suicide rate increases with age. (2.4)
This remains true when controlling for continent in the Americas, Asia & Europe, but not for Africa & Oceania. (2.8)
There is a weak positive relationship between a countries GDP (per capita) and suicide rate. (2.10)
The highest suicide rate ever recorded in a demographic (for 1 year) is 225 (per 100k population). (2.12)
There is an overrepresentation of men in suicide deaths at every level of analysis (globally, at a continent and country level). Globally, the male rate is ~3.5x higher. (2.3) (2.6) (2.7)
```{r}
# 2.1 Global Trend
data %>%
  group_by(year) %>%
  summarize(population = sum(population), 
            suicides = sum(suicides_no), 
            suicides_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = year, y = suicides_per_100k)) + 
  geom_line(col = "deepskyblue3", size = 1) + 
  geom_point(col = "deepskyblue3", size = 2) + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey35", size = 1) +
  labs(title = "Global Suicides (per 100k)",
       subtitle = "Trend over time, 1985 - 2015.",
       x = "Year", 
       y = "Suicides per 100k") + 
  scale_x_continuous(breaks = seq(1985, 2015, 2)) + 
  scale_y_continuous(breaks = seq(10, 20))

ggsave("2_1.png", width = 8, height = 6, device = "png", dpi = 300)

#Insights

#Peak suicide rate was 15.3 deaths per 100k in 1995
#Decreased steadily, to 11.5 per 100k in 2015 (~25% decrease)
#Rates are only now returning to their pre-90’s rates
#Limited data in the 1980’s, so it’s hard to say if rate then was truly representative of the global population
```

```{r}
# 2.2 By Continent

continent <- data %>%
  group_by(continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  arrange(suicide_per_100k)

continent$continent <- factor(continent$continent, ordered = T, levels = continent$continent)

continent_plot <- ggplot(continent, aes(x = continent, y = suicide_per_100k, fill = continent)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Global Suicides (per 100k), by Continent",
  x = "Continent", 
  y = "Suicides per 100k", 
  fill = "Continent") +
  theme(legend.position = "none", title = element_text(size = 10)) + 
  scale_y_continuous(breaks = seq(0, 20, 1), minor_breaks = F)


continent_time <- data %>%
  group_by(year, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

continent_time$continent <- factor(continent_time$continent, ordered = T, levels = continent$continent)

continent_time_plot <- ggplot(continent_time, aes(x = year, y = suicide_per_100k, col = factor(continent))) + 
  facet_grid(continent ~ ., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Continent", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Continent") + 
  theme(legend.position = "none", title = element_text(size = 10)) + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)

g <- grid.arrange(continent_plot, continent_time_plot, ncol = 2)

ggsave("2_2.png", g, width = 8, height = 6, device = "png", dpi = 300)
```
























